Parallel Global Search Algorithm for Optimization of the Kinetic Parameters of Chemical Reactions

### 2021-04-15 2021/RSD_Ufa/RSD2021.tex
new figures from Ufa

\usepackage[russian]{babel}

@@

\begin{document}
\mainmatter              % start of a contribution
%
\title{Parallel Global Search Algorithm for Optimization of the Kinetic Parameters of Chemical Reactions}
%
\titlerunning{Adaptive Global Optimization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{
Irek Gubaydullin$^{1,2}$\and
Leniza Enikeeva$^{2,3}$\orcidID{0000-0003-4219-4870}
\and
Konstantin Barkalov$^4$\orcidID{0000-0001-5273-2471}
\and
Ilya Lebedev$^4$\orcidID{0000-0002-8736-0652} 
}

@@

\begin{abstract}
The paper considers the application of parallel computing technology to the simulation of a catalytic chemical reaction, which is widely used in the modern chemical industry to produce synthesis gas. As a chemical reaction, the process of pre-reforming propane on a Ni catalyst is assumed. To simulate a chemical process, it is necessary to develop a kinetic model of the process, that is, to determine the kinetic parameters. To do this, the inverse problem of chemical kinetics is solved, which predicts the values of kinetic parameters based on laboratory data. From a mathematical point of view, the inverse of the chemical kinetics problem is a global optimization problem. A parallel information-statistical global search algorithm was used to solve it. The use of the parallel algorithm has reduced significantly the search time for the optimum. The found optimal parameters of the model made it possible to adequately simulate the process of pre-reforming propane on a Ni-catalyst.

\keywords{Global optimization $\cdot$ Multiextremal functions $\cdot$ Parallel computing }
\end{abstract}

@@

One of the areas of the Laboratory of Mathematical Chemistry of the Institute of Petrochemistry and Catalysis of the UFIC RAS is the study of chemical reactions by mathematical modeling methods. At the moment, a large number of complex chemical processes have been studied-the reaction of cycloalumination of alkenes with triethylaluminium into alumacyclopentanes, the reaction of hydroalumination of olefins with organoaluminum compounds, the reaction of alcohols with dimethyl carbonate, etc. The mathematical description of chemical reactions is necessary for the construction of mathematical models of chemical kinetics, which are called kinetic models. The kinetic model is the basis for mathematical modeling of chemical reactions and the initial level for mathematical modeling of chemical reactors. When developing a kinetic model of a multi-stage chemical reaction, the inverse problem of chemical kinetics (CK) is solved. The inverse problem of HC is generally called the process of selecting the mechanism of a complex process and determining its quantitative characteristics - the rate constants of single-stage reactions based on experimental data on the dynamics of the process. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem, so the problem of choosing an optimization method is relevant.

@@

The features of inverse problems of chemical kinetics listed above were taken into account in the parallel global search algorithm developed in University of Nizhni Novgorod. The information-statistical approach described in \cite{Strongin2000} in details makes a theoretical basis for this algorithm. The developed algorithm is based on a reduction of the initial multidimensional problem to an equivalent one-dimensional problem followed by solving the latter by efficient optimization methods for the functions of single variable. The issues of the parallelization of the algorithm on various architectures were considered in \cite{Barkalov2016}. In the present paper, the results obtained with the use of the parallel global search algorithm for solving the inverse problem of chemical kinetics are presented.

@@

The process under study is propane pre-reforming into methane-rich gas over Ni catalyst, which is an industrially important chemical process \cite{USKOV2019126741, Uskov_en}. Pre-reforming of propane was studied over industrial nickel-chromium catalyst at pressure of 1 and 5 bar in the temperature range of 220 -- 380 $^0$C and at flow rates of 4000 and 12000 h$^{-1}$. The experimental data on propane pre-reforming was acquired from our previous work \cite{Uskov_en}.  The reaction scheme consists of two reactions: propane steam conversion and CO$_2$ methanation \cite{Uskov_cat}:
\begin{gather}
C_3H_8 + 6H_2O \rightarrow 10H_2 + 3CO_2 \label{reac_ref}\\
CO_2 + 4H_2 \rightleftharpoons CH_4 + 2H_2O \label{reac_met},
\end{gather}

where the rates of reactions \eqref{reac_ref} -- \eqref{reac_met} are expressed according to the Langmuir-Hinshelwood model \cite{USKOV2019126741}:

@@

\[
W_{met} = k_{met}^0 \cdot \exp \left(- \dfrac{E_{met}}{RT} \right) \cdot C_{H2} \cdot \left[1 - \dfrac{P_{CH4} P_{H2O}^2}{K_{eq} P_{CO2} P_{H2}^4 }\right] ,
\]

where $W_{ref}$ and $W_{met}$ are the reaction rates; $E_{ref}$ and $E_{met}$ are the observed activation energies, J/mol; $k^0_{ref}$ and $k^0_{met}$ are the pre–exponential multipliers; $B$ is the constant parameter, $T$ is the temperature, K; $R$ is the universal gas constant, J(K$\cdot$ mol). The "$ref$"and "$met$" indexes refer to pre-reforming and methanation reactions, respectively. $C_{C3H8}$ and $C_{H2}$ are concentrations of propane and hydrogen, mol/m$^3$; $m$ is order of the denominator, which varied from 0 to 2; $K_{eq}$ is the equilibrium constant of $CO_2$ methanation; $P_{CH4}$, $P_{H2O}$, $P_{CO2}$, $P_{H2}$ are partial pressures of the corresponding substances, bar. The mathematical model is a system of equations of material balance:

\begin{equation}\label{mater_balance}
\begin{cases}
G \dfrac{dy_i}{dl} = \left(\nu_i^{ref} W_{ref} + \nu_i^{met} W_{met} \right) m_i, 
\\
0 \le l \le , i \in \left\lbrace C_3H_8, CH_4, H_2O, H_2, CO_2 \right\rbrace,
\\
l = 0, y_i = y_{i0},
\end{cases}
\end{equation}

where $G$ is a mass flow of the mixture, kg/(m$^2 \cdot$ sec); $y_i$ is a mass fraction of the $i$-th component; $\nu_i$ is a stoichiometric coefficient of the $i$-th component; $m_i$ is a molar mass of the $i$-th component, kg/mol; $l$ is coordinate along the catalytic layer, m; $L$ is a length of the catalytic layer, m. The length of the catalytic layer is 0.008 m. The mathematical model of chemical kinetics problems is a system of differential equations that describes the variations in substance concentrations over time according to the rates of reaction stages. The system of differential equations is a Cauchy problem containing the initial data \cite{ENIKEEV2020123131, Chainikova}. The numerical solving of such a system of equations is a direct problem of chemical kinetics.

@@

Systems of ordinary differential equations (ODES) describing chemical processes are often rigid due to the presence of fast-flowing and slow-flowing reaction stages. Therefore, the RADAU-IIA method was chosen as a method for solving the ODE system, which is suitable for solving rigid ODE systems. Let us consider in the general case the Cauchy problem of the form:

\begin{equation}\label{ode}
\dfrac{dx}{dt} = f(\textbf{x}, y), \textbf{x}(0) = \textbf{x}_0, 0 \le t \le T,
\end{equation}

%где x в нашем случае это вектор концентраций веществ реакции размерности 5, а f это правая часть уравнения \eqref{mater_balance}. Тогда s-стадийный метод Рунге-Кутты для решения системы \eqref{ode} выписывается следующим образом:

where $x$ in our case is the vector of concentrations of reaction substances of dimension 5, and $f$ is the right-hand side of the equation \eqref{mater_balance}. Then the $s$-stage Runge-Kutta method for solving the \eqref{ode} system is written as follows:

\begin{equation}\label{runge}
\begin{cases}
X_i = x_n + h \sum\limits_{j=1}^s a_{ij} f(X_j, t_n + c_j h) , 1 \le i \le s, 
\\
x_{n+1} = x_n + h \sum\limits_{i=1}^s b_i f (X_i, t_n + c_i h),
\end{cases}
\end{equation}

%где $x_n$, $x_{n+1}$, $X_i$ -- приближенные решения системы \eqref{ode} в моменты времени $t_n$, $t_{n+1} = t_n + h$, $t_n + c_i h$. Коэффициенты  методов  Рунге-Кутты  обычно  записывают  в  виде таблицы Бутчера:

where $x_n$, $x_{n+1}$, $X_i$ are approximate solutions of the \eqref{ode} system at time points $t_n$, $t_{n+1} = t_n + h$, $t_n + c_i h$. The coefficients of the Runge-Kutta methods are usually written in the form of a Butcher table:

@@

The Butcher table contains the vectors $b$ and $c$, consisting of the coefficients $b_i$ and $c_i$, respectively, and the matrix $A$ of dimension $s \times s$, including the coefficients $a_{ij}$.

Fifth-order Rado-IIA method (Radau 5):

@@

Determining the kinetic parameters of reaction stages by comparing calculated values of substance concentrations and experimental results is an inverse problem of chemical kinetics. The mathematical problem is to minimize the functional of the deviation between calculated and experimental values. The functional of minimizations determined as the sum of absolute deviations between calculated and experimental concentrations:

\begin{equation}\label{func}
F = \sum\limits_{i=1}^M \sum\limits_{j=1}^N \left| x_{ij}^{calc} - x_{ij}^{exp} \right| \rightarrow \min,
\end{equation}

where $x_{ij}^{calc}$ and $x_{ij}^{exp}$ are calculated and experimental values of component concentrations; $M$ is the number of measuring points; $N$ is the number of substances involved in the reaction.

@@

\section{Numerical Experiments}\label{Sec_Exp}

@@

\begin{figure}[ht]
\center{\includegraphics[width=0.8\linewidth]{time.pdf}}
\caption{The parallrl algorithm running time}
\label{fig_time}
\end{figure}

@@

@@

\textbf{Acknowledgments}. This study was supported by the Russian Science Foundation, project No.\,21-11-00204.

>>>>

%\usepackage[russian]{babel}

@@

\begin{document}
\mainmatter              % start of a contribution
%
\title{Parallel Global Search Algorithm for Optimization of the Kinetic Parameters of Chemical Reactions}
%
\titlerunning{Parallel Global Search}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{
Irek Gubaydullin$^{1,2}$\and
Leniza Enikeeva$^{2,3}$\orcidID{0000-0003-4219-4870}
\and
Konstantin Barkalov$^4$\orcidID{0000-0001-5273-2471}
\and
Ilya Lebedev$^4$\orcidID{0000-0002-8736-0652} 
}

@@

\begin{abstract}
The paper considers the application of parallel computing technology to the simulation of a catalytic chemical reaction, which is widely used in the modern chemical industry to produce synthesis gas. As a chemical reaction, the process of pre-reforming propane on a Ni catalyst is assumed. To simulate a chemical process, it is necessary to develop a kinetic model of the process, that is, to determine the kinetic parameters. To do this, the inverse problem of chemical kinetics is solved, which predicts the values of kinetic parameters based on laboratory data. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem. A parallel information-statistical global search algorithm was used to solve it. The use of the parallel algorithm has significantly reduced the search time to find the optimum. The found optimal parameters of the model made it possible to adequately simulate the process of pre-reforming propane on a Ni-catalyst.

\keywords{Global optimization $\cdot$ Multiextremal functions $\cdot$ Parallel computing $\cdot$ Chemical kinetics $\cdot$ Inverse problems }
\end{abstract}

@@

One of the areas of the Laboratory of Mathematical Chemistry of the Institute of Petrochemistry and Catalysis of the UFIC RAS is the study of chemical reactions by mathematical modeling methods. At the moment, a large number of complex chemical processes have been studied, among  them, the reaction of cycloalumination of alkenes with triethylaluminium into alumacyclopentanes, the reaction of hydroalumination of olefins with organoaluminum compounds, the reaction of alcohols with dimethyl carbonate, etc. The mathematical description of chemical reactions is necessary for the development the mathematical models of chemical kinetics, which are called kinetic models. The kinetic model is the basis for mathematical modeling of chemical reactions and the initial level for mathematical modeling of chemical reactors. When developing a kinetic model of a multi-stage chemical reaction, the inverse problem of chemical kinetics (CK) is solved. The inverse problem of CK is generally called the process of selecting the mechanism of a complex process and determining its quantitative characteristics -- the rate constants of reactions based on experimental data about the dynamics of the process. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem, so the problem of choosing an optimization method is relevant.

@@

The features of inverse problems of chemical kinetics listed above were taken into account in the parallel global search algorithm developed in University of Nizhni Novgorod. The information-statistical approach described in \cite{Strongin2000} in details makes a theoretical basis for this algorithm. The developed algorithm is based on a reduction of the initial multidimensional problem to an equivalent one-dimensional problem followed by solving the latter by efficient optimization methods for the functions of single variable. The issues of the parallelization of the algorithm on various architectures were considered in \cite{Barkalov2016,Strongin2018,Barkalov2020}. In the present paper, the results obtained with the use of the parallel global search algorithm for solving the inverse problem of chemical kinetics are presented.

@@

The process under study is propane pre-reforming into methane-rich gas over Ni catalyst, which is an industrially important chemical process \cite{USKOV2019126741, Uskov_en}. Pre-reforming of propane was studied over industrial nickel-chromium catalyst at pressure of 1 and 5 bar in the temperature range of 220 -- 380 $^0$C and at flow rates of 4000 and 12000 h$^{-1}$. The experimental data on propane pre-reforming was acquired from our previous work \cite{Uskov_en}.  The reaction scheme consists of two reactions: propane steam conversion and CO$_2$ methanation \cite{Uskov_cat}:
\begin{gather}
C_3H_8 + 6H_2O \rightarrow 10H_2 + 3CO_2 \label{reac_ref}\\
CO_2 + 4H_2 \rightleftharpoons CH_4 + 2H_2O \label{reac_met},
\end{gather}
where the rates of reactions \eqref{reac_ref} -- \eqref{reac_met} are expressed according to the Langmuir-Hinshelwood model \cite{USKOV2019126741}:

@@

\[
W_{met} = k_{met}^0 \cdot \exp \left(- \dfrac{E_{met}}{RT} \right) \cdot C_{H2} \cdot \left[1 - \dfrac{P_{CH4} P_{H2O}^2}{K_{eq} P_{CO2} P_{H2}^4 }\right] ,
\]
where $W_{ref}$ and $W_{met}$ are the reaction rates; $E_{ref}$ and $E_{met}$ are the observed activation energies, J/mol; $k^0_{ref}$ and $k^0_{met}$ are the pre–exponential multipliers; $B$ is the constant parameter, $T$ is the temperature, K; $R$ is the universal gas constant, J/(K$\cdot$ mol). The ``$ref$''and ``$met$'' indexes refer to pre-reforming and methanation reactions, respectively. $C_{C3H8}$ and $C_{H2}$ are concentrations of propane and hydrogen, mol/m$^3$; $m$ is order of the denominator, which varied from 0 to 2; $K_{eq}$ is the equilibrium constant of $CO_2$ methanation; $P_{CH4}$, $P_{H2O}$, $P_{CO2}$, $P_{H2}$ are partial pressures of the corresponding substances, bar. The mathematical model is a system of equations of material balance:

\begin{equation}\label{mater_balance}
\begin{cases}
G \dfrac{dy_i}{dl} = \left(\nu_i^{ref} W_{ref} + \nu_i^{met} W_{met} \right) m_i, 
\\
0 \le l \le , i \in \left\lbrace C_3H_8, CH_4, H_2O, H_2, CO_2 \right\rbrace,
\\
l = 0, y_i = y_{i0},
\end{cases}
\end{equation}
where $G$ is a mass flow of the mixture, kg/(m$^2 \cdot$ sec); $y_i$ is a mass fraction of the $i$-th component; $\nu_i$ is a stoichiometric coefficient of the $i$-th component; $m_i$ is a molar mass of the $i$-th component, kg/mol; $l$ is coordinate along the catalytic layer, m; $L$ is a length of the catalytic layer, m. The length of the catalytic layer is 0.008 m. The mathematical model of chemical kinetics problems is a system of differential equations that describes the variations in substance concentrations over time according to the rates of reaction stages. The system of differential equations is a Cauchy problem containing the initial data \cite{ENIKEEV2020123131, Chainikova}. The numerical solving of such a system of equations is a direct problem of chemical kinetics.

@@

Systems of ordinary differential equations (ODES) describing chemical processes are often stiff due to the presence of fast-flowing and slow-flowing reactions. Therefore, the RADAU-IIA method was chosen as a method for solving the ODE system, which is suitable for solving stiff ODE systems. Let us consider in the general case the Cauchy problem of the form:

\begin{equation}\label{ode}
\dfrac{dx}{dt} = f(\textbf{x}, y), \textbf{x}(0) = \textbf{x}_0, 0 \le t \le T,
\end{equation}
%где x в нашем случае это вектор концентраций веществ реакции размерности 5, а f это правая часть уравнения \eqref{mater_balance}. Тогда s-стадийный метод Рунге-Кутты для решения системы \eqref{ode} выписывается следующим образом:
where $x$ is the vector of concentrations of reaction substances with dimension of 5 (the substances are $C_3H_8$, $CH_4$, $H_2O$, $H_2$, $CO_2$), and $f$ is the right-hand side of the equation \eqref{mater_balance}. Then the $s$-stage Runge-Kutta method for solving the \eqref{ode} system is written as follows:

\begin{equation}\label{runge}
\begin{cases}
X_i = x_n + h \sum\limits_{j=1}^s a_{ij} f(X_j, t_n + c_j h) , 1 \le i \le s, 
\\
x_{n+1} = x_n + h \sum\limits_{i=1}^s b_i f (X_i, t_n + c_i h),
\end{cases}
\end{equation}
%где $x_n$, $x_{n+1}$, $X_i$ -- приближенные решения системы \eqref{ode} в моменты времени $t_n$, $t_{n+1} = t_n + h$, $t_n + c_i h$. Коэффициенты  методов  Рунге-Кутты  обычно  записывают  в  виде таблицы Бутчера:  
where $x_n$, $x_{n+1}$, $X_i$ are approximate solutions of the system \eqref{ode} at time points $t_n$, $t_{n+1} = t_n + h$, $t_n + c_i h$. The coefficients of the Runge-Kutta methods are usually written in the form of a Butcher tableau:

@@

The Butcher tableau contains the vectors $b$ and $c$, consisting of the coefficients $b_i$ and $c_i$, respectively, and the matrix $A$ of dimension $s \times s$, including the coefficients $a_{ij}$. The coefficients of fifth-order Rado-IIA method (Radau 5) are presented below:

@@

The solution of the system of ordinary differential equations is a vector of the calculated concentrations of the reaction components, which are then compared with the experimental data.
Determining the kinetic parameters of reaction stages by comparing calculated values of substance concentrations and experimental results is an inverse problem of chemical kinetics. The mathematical problem is to minimize the functional of the deviation between calculated and experimental values. The objective function determined as the sum of absolute deviations between calculated and experimental concentrations:

\begin{equation}\label{func}
F = \sum\limits_{i=1}^M \sum\limits_{j=1}^N \left| x_{ij}^{calc} - x_{ij}^{exp} \right| \rightarrow \min,
\end{equation}
where $x_{ij}^{calc}$ and $x_{ij}^{exp}$ are calculated and experimental values of component concentrations; $M$ is the number of measuring points; $N$ is the number of substances involved in the reaction.

@@

\section{Numerical Experiments}\label{Sec_Exp}
%Таким образом, в результате решения обратной задачи химической кинетики параллельным методом глобального поиска были рассчитаны кинетические параметры процесса предриформинга пропана на Ni катализаторе (Table \ref{tab:1}). Для интерпретации полученных результатов проводится решение прямой задачи химической кинетики, то есть расчет концентраций компонентов реакции и сравнение их с экспериментальными данными. На рисунках НОМЕР a, b показано сравнение рассчитанных концентраций с экспериментальными при разных условиях: (a) GHSV 4000 h$^{-1}$, (b) GHSV 12000 h$^{-1}$. Результаты расчетов качественно согласуются с экспериментом. Ошибка отклонения составляет не более ПОСЧИТАТЬ \%.
Thus, as a result of solving the inverse problem of chemical kinetics by the parallel global search method, the kinetic parameters of the propane pre-reforming process on a Ni catalyst were calculated (Table \ref{tab:1}). To interpret the results obtained, the solution of the direct problem of chemical kinetics was carried out, namely, the concentrations of the reaction components were calculated and compared with the experimental data. Fig.~\ref{fig_kin_curves} shows a comparison of the calculated concentrations with the experimental ones under different conditions: (a) GHSV=4000 h$^{-1}$, (b) GHSV=12000 h$^{-1}$. The results of the calculations are qualitatively consistent with the experiment.

\begin{figure}[ht]
	\begin{minipage}{0.5\linewidth}
		\center{\includegraphics[width=1.0\linewidth]{kin_curve1.png} \\ (a)}
	\end{minipage}
	\begin{minipage}{0.5\linewidth}
		\center{\includegraphics[width=1.0\linewidth]{kin_curve2.png} \\ (b)}
	\end{minipage}
	\caption{Temperature dependences of the output concentrations of propane C3H8, methane CH4, hydrogen H2 and CO2 in the process of propane pre-reforming. Experimental conditions: 220-380 $^0$C, (a) GHSV = 4000 h$^{-1}$, (b) GHSV = 12000 h$^{-1}$, 1 bar pressure. Concentrations of the gas components on the figure are given on the dry basis. Points are experimental concentrations (``exp''-index), lines are simulated concentrations (``calc''-index).}
	\label{fig_kin_curves}
\end{figure}

\begin{table}[ht]
	\caption{Calculated kinetic parameters}
	\label{tab:1}
	\center
	\begin{tabular}{c|c|c|c|c|c}
		\hline
		%\noalign{\smallskip}
		$E_{ref}$, kJ/mol & $E_{met}$, kJ/mol & $k_{ref}$, (mole$\cdot$m$^{-3}$)m$\cdot$s$^{-1}$ & $k_{met}$, s$^{-1}$ & $m$ & $B$ \\
		%\noalign{\smallskip} 
		\hline 
		%\noalign{\smallskip}
		99.23 \;&	37.08 \; & $2.4 \cdot 10^{10}$ \;  & $1.93 \cdot 10^5$ \; & 0.82 & 4.8  \\
		%\noalign{\smallskip}
		\hline
	\end{tabular}
\end{table}
%Для сравнения с предыдущими расчетами был построен график арренисовской зависимости (рисунок НОМЕР), на котором представлены зависимости десятичного логарифма от констант скоростей реакций \eqref{reac_ref} и \eqref{reac_met} от обратной температуры. Обозначения "GA", "GSA" и "PGSA" означают  решения, полученные различными алгоритмами оптимизации, соотвественно, генетическим алгоритмом, алгоритмом гравитационного поиска и параллельным алгоритмом глобального поиска.

For comparison with the previous calculations, an Arrhenis plot was constructed (Fig.~\ref{fig_arren}), which shows the dependence of the decimal logarithm on the reaction rate constants \eqref{reac_ref} and \eqref{reac_met} on the inverse temperature. The designations ``GA'', ``GSA'' and ``PGSA'' mean solutions obtained by various optimization algorithms, the genetic algorithm, the gravity search algorithm, and the parallel global search algorithm, respectively. The blue lines refer to the pre-reforming reaction \eqref{reac_ref}, and the yellow lines refer to the methanation reaction~\eqref{reac_met}. Note that the lines of the same color are parallel to each other, which means that the activation energies of the reaction calculated by different methods have approximately the same values, which indicates the adequacy of the calculations.

\begin{figure}[ht]
	\center{\includegraphics[width=0.8\linewidth]{arren.png}}
	\caption{Comparison of the obtained solution with the other algorithms found on the Arrhenius plot}
	\label{fig_arren}
\end{figure}

@@

\begin{figure}[ht]
\center{\includegraphics[width=0.8\linewidth]{time.pdf}}
\caption{The parallel algorithm running time}
\label{fig_time}
\end{figure}

@@

%В будущем планируется моделирование других химических процессов, а именно, первая задача это моделирование процесса сернокислотного алкилирования изобутана олефинами, для которого на данный момент составлена схема химических превращений, проведен литературный обзор процесса и найдены экспериментальные данные. Математическая модель реакции сернокислотного алкилирования изобутана олефинами представляет собой задачу Коши, на данный момент решена прямая задача химической кинетики, далее будет решаться обратная задача предложенными параллельными методами. Вторая задача это моделирование процесса синтеза метил-трет бутилового эфира. Химизм данного процесса известен, задача иемют небольшую размерность, однако моделирование реактора трудоемко и представляет собой решение уравнение в частных производных второго порядка. На данный момент расписана разностная схема, реализован метод прогонки, следующим этапом будет решение прямой задачи химической кинетики, а также предсказание кинетических параметров процесса предложенными параллельными методами.

In the future, it is planned to simulate other chemical processes, namely, the first task is to simulate the process of sulfuric acid alkylation of isobutane with olefins, for which a scheme of chemical transformations has been compiled, a literature review of the process has been conducted, and experimental data have been found. The mathematical model of the reaction of sulfuric acid alkylation of isobutane by olefins is a Cauchy problem, at the moment the direct problem of chemical kinetics is solved, then the inverse problem will be solved by the proposed parallel methods. The second task is to simulate the synthesis of methyl-tert-butyl ether. The chemistry of this process is known, and the problem has a small dimension, but the simulation of the reactor is time-consuming and represents the solution of a second-order partial differential equation. At the moment, the difference scheme is formed, the run-through method is implemented, the next stage will be the solution of the direct problem of chemical kinetics, as well as the prediction of the kinetic parameters of the process by the proposed parallel methods.

@@

\textbf{Acknowledgments}. This study was supported by the Russian Science Foundation, project No.\,21-11-00204 and by RFBR, project No.\,19-37-60014.
